Type theories as quotient inductive-recursive types

### 2025-02-08 tex/main.tex
Minor revisions; fix a typo in S4.2

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B(x)$.
%
We are further inspired by Agda in writing $\implicit{x}B(x)$ for $(x : A) \to B(x)$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.

@@

We have formally implemented our definitions and proofs in Agda, with the following options: \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}. We simulate quotient inductive types using postulated equations to introduce equality constructors~\cite{Licata2011}.
Some of our theorem statements are meta-theoretical, and thus cannot be checked by Agda. Others can only be checked  partially, because of postulated quotient inductive types. We have marked such ``informal'' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume uniqueness of identity proofs globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{X}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil & : \JList{X} \\
  \Jsing{\blank} & : X \to \JList{X} \\
  \Jconcat{\blank}{\blank} & : \JList{X} \to \JList{X} \to \JList{X} \\
  \mathsf{idl} & : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{X}} \ell \\
  \mathsf{idr} & : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{X}} \ell \\
  \mathsf{assoc} & : \implicit{\ell_0, \ell_1, \ell_2}\; (\Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2}) =^{\JList{X}} (\Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})})
\end{alignat*}

@@

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: The type $\JList{X}$ is defined simultaneously with a recursively defined function $\flatten : \JList{X} \to \List{X}$ which flattens a join list into an ordinary list. We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
\mathsf{nf} : (\ell, \ell' : \JList{X}) \to \flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell' \to \ell =^{\JList{X}} \ell'
\]
which states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{X}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,X} \flatten\,\ell'$.

@@

\LT{How the congruence rule for the identity type can be used in QII(R)T?}

@@

We begin with the definition of the parallel substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning.
Then, we develop a type theory with a universe (\cref{subsec:SC+U}), $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).

@@

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ are kept implicit.
\Fredrik{Could we change this to $A[\sigma]$ instead?}
\LT{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition: }

@@

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which are intended to substitute the most recently bound variable by the term~$t$, and act like $\sigma$ on the other variables:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$, because the substitution calculus is supposed to be the initial \emph{category} with families:
%Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families),
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
\end{alignat*}
These need to satisfy certain laws, which we introduce later as equality constructors.
Context extension is characterised by its projections, which picks out the rest of the context and the most recent variable
%Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution, which given a substitution $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in context $\Delta$ to terms in context $\Gamma$, appropriately preserving the type, which is also substituted:
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A).
\end{alignat*}
Note that $[\blank]\blank$ is overloaded for type and term substitution.

The structural rules for type substitution are stipulated by following equality constructors:
\begin{alignat*}{3}
  [\idS]_T & : \implicit{\Gamma, A}                               && [ \idS ] \;A         && =^{\Ty\,\Gamma}\;A \\
  [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\;A && =^{\Ty\,\Gamma} [ \sigma ]\;([ \tau ]\;A) \\
  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                && [ \sigma ]\;\UU        && =^{\Ty\,\Gamma} \UU
\end{alignat*}

@@

The laws for substitution composition are stipulated using equality constructors, where, for the first time, we see the need to adjust the types involved with a $\transp$ in order for both sides to have the same type.
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
\end{alignat*}
%
Composition of substitutions $\blank;\blank$ acts like a list concatenation, so
we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), [\sigma]\,t$. However $[\sigma]\,t$ is of type $\Tm\;\Gamma\;([\sigma]\,[\tau]\;A)$, but the context extension $(\sigma; \tau), [\sigma]\,t$ requires $[\sigma]\,t$ to have type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead. Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ which we can transport along to fix up the equation.

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & \defeq A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & \defeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will only fire if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
where we have written $t \reduce u$ for $t \equiv u$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

To repair the local confluence, we add a simultaneously defined function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
\begin{alignat*}{3}
  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
[ \idS ]_{\Tm}\,t          & \reduce t \\
[ \sigma ; \tau ]_{\Tm}\,t & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
[ \pi_1(\sigma, t) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\,t \\
[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
[ \sigma ]_{\Tm}\,t        & \reduce [ \sigma ]\,t, \quad \text{otherwise}
\end{alignat*}
which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
\begin{remark}
  The function $[\blank]_{\Tm}\blank$ can be understood as a \emph{smart constructor} for the explicit term substitution $[\blank]\blank$, a well-known idiom in functional programming, since the function $[\blank]_{\Tm}\blank$ (mostly) builds an explicit substitution but computes certain cases away.
\end{remark}

@@

Similar to term substitution, we can show that the lifted substitution $\sigma \uparrow A$ is propositionally equal to $\sigma^+$, based on two basic properties\footnote{%
  The reader is invited to prove them using the QII definition of type theory.}
about the lifting.
\begin{lemma} \label{lem:lifting}
  The following statements hold:
  \[
    \idS^+ =^{\Sub\;(\Gamma, A)\;(\Gamma, A)} \idS
    \qquad\text{and}\qquad
    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, [\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;(\Theta, A)} \sigma^+ ; \tau^+.
  \]
\end{lemma}

>>>>

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B(x)$.
%
We are further inspired by Agda in writing $\implicit{x}B(x)$ for $(x : A) \to B(x)$ when $x$ is treated as an ``implicit'' argument which is not given explicitly, but instead can be figured out from context.
\LT{We write $\Set$ for the type of types employing the `typical ambiguity'.}

@@

We have formally implemented our definitions and proofs in Agda, with the following options: \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}. We simulate quotient inductive types using postulated equations to introduce equality constructors~\cite{Licata2011}.
Some of our theorem statements are meta-theoretical, and thus cannot be checked by Agda. Others can only be checked  partially, because of postulated quotient inductive types. We have marked such ``informal'' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.
\LT{Single quote or double quote in British English?}

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume uniqueness of identity proofs globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}

@@

We can also define join lists as a quotient inductive-\emph{recursive} type, as follows: The type $\JList{A}$ is defined simultaneously with a recursion $\flatten : \implicit{A}\;\JList{A} \to \List{A}$ which flattens a join list into an ordinary list. We keep the constructors $\Jnil$, $\Jsing{\blank}$ and $\Jconcat{\blank}{\blank}$, but replace $\mathsf{idl}$, $\mathsf{idr}$ and $\mathsf{assoc}$ by a single constructor
\[
  \mathsf{nf} : \implicit{A}\;(\ell, \ell' : \JList{A}) \to \flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell' \to \ell =^{\JList{A}} \ell'
\]
which states that join lists are equal as soon as their flattenings are equal. For $\flatten$ to be well defined, it needs to respect the $\mathsf{nf}$ constructor, but this is a tautology: if $\ell =^{\JList{A}} \ell'$ because $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$, then obviously $\flatten\,\ell =^{\mathsf{List}\,A} \flatten\,\ell'$.

@@

@@

We begin with the definition of the parallel substitution calculus~\cite{Martin-Lof1992} as a QIIT, emphasising how the use of transports in its very definition complicates formal reasoning.
We then introduce its QIIRT counterpart (\cref{subsec:SC-QIIRT}) which shortens the gap between formal and informal reasoning, and develop a type theory with a universe (\cref{subsec:SC+U}) of small types, $\Pi$-types (\cref{subsec:SC+U+Pi}), and other type formers (\cref{subsec:SC+U+Pi+more}).

@@

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ are kept implicit.
\Fredrik{Could we change this to $A[\sigma]$ instead?}
\LT{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition}
\LT{The rule \textsf{concat} doesn't look like concatenation with the $\tau \comp \sigma$ because the $\sigma$ is a snoc-list.}

@@

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as (snoc-)lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which are intended to substitute the most recently bound variable by the term~$t$, and act like $\sigma$ on the other variables:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;                && \Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\,A$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$, because the substitution calculus is supposed to be the initial \emph{category} with families:
%Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families),
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;                         && \Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;&& \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
\end{alignat*}
These need to satisfy certain laws, which we introduce later as equality constructors.
Context extension is characterised by its projections, which picks out the rest of the context and the most recent variable
%Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\, A)
\end{alignat*}
respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution, which given a substitution $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in context $\Delta$ to terms in context $\Gamma$, appropriately preserving the type, which is also substituted:
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\, A).
\end{alignat*}
Note that $[\blank]\blank$ is overloaded for type and term substitution.

The structural rules for type substitution are stipulated by following equality constructors:
\begin{alignat*}{3}
  [\idS]_T & : \implicit{\Gamma, A}                               && [ \idS ] \,A         && =^{\Ty\,\Gamma}\;A \\
  [;]_T    & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A} && [ \sigma ; \tau ]\,A && =^{\Ty\,\Gamma} [ \sigma ]\,([ \tau ]\,A) \\
  []\UU      & : \implicit{\Gamma, \Delta, \sigma}                && [ \sigma ]\,\UU        && =^{\Ty\,\Gamma} \UU
\end{alignat*}

@@

The laws for substitution composition are stipulated using equality constructors, where, for the first time, we see the need to adjust the types involved with a $\transp$ in order for both sides to have the same type.
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
\end{alignat*}
%
Composition of substitutions $\blank;\blank$ acts like a list concatenation, so
we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), [\sigma]\,t$. However $[\sigma]\,t$ is of type $\Tm\;\Gamma\;([\sigma]\,[\tau]\;A)$, but the extension $(\sigma; \tau), [\sigma]\,t$ requires $[\sigma]\,t$ to have type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead. Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ which we can transport along to fix up the equation.

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & \defeq A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & \defeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
where we have written $t \reduce u$ for $t \equiv u$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

To repair the local confluence, we add a simultaneously defined function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
\begin{alignat*}{3}
  [\blank]_{\Tm}\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
[ \idS ]_{\Tm}\,t          & \reduce t \\
[ \sigma ; \tau ]_{\Tm}\,t & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
[ \pi_1(\sigma, t) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\,t \\
[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
[ \sigma ]_{\Tm}\,t        & \reduce [ \sigma ]\,t, \quad \text{otherwise}
\end{alignat*}
which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
\begin{remark}
  The function $[\blank]_{\Tm}\blank$ can be understood as a \emph{smart constructor} for the explicit term substitution $[\blank]\blank$, a well-known idiom in functional programming, since the function $[\blank]_{\Tm}\blank$ (mostly) builds an explicit substitution but computes certain cases away.
\end{remark}

@@

Similar to term substitution, we can show that the lifted substitution $\sigma \uparrow A$ is propositionally equal to $\sigma^+$, based on two basic properties\footnote{%
  The reader is invited to prove them using the QII definition with transports.}
about the lifting.
\begin{lemma} \label{lem:lifting}
  The following statements hold:
  \[
    \idS^+ =^{\Sub\;(\Gamma, A)\;(\Gamma, A)} \idS
    \qquad\text{and}\qquad
    (\sigma; \tau)^+     =^{\Sub\;(\Gamma, [\sigma; \tau]_{\Ty}\,A)\;(\Theta, A)} \sigma^+ ; \tau^+.
  \]
\end{lemma}
